---
title: "Boote et al. 2022"
output: html_notebook
---

This notebook is related to the DE analysis for Boote et al 2022
The input is a set of abundance tables from running Kallisto
that are read in to run a DESEQ2 analysis.

There is an associated meta info file boote_meta_info.csv. NB that I have spent some
time cleaning up this .csv so use this one rather than some other.

See the nextflow pipeline for details of the preprocessing that has been performed
and details of the Kallisto run to create count tables.

There are two species: Acropora intermedia and Montipora digitata
For each species there are three treatments Deoxy, Normoxy and Time_zero
As a start this script will run pairwise comparisons of each of the treatments within each of the species

I have output results with and without using schrinkage
See here for further details: # see here: http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#log-fold-change-shrinkage-for-visualization-and-ranking

Load required packages.
```{r}
library(tximport)
library(readr)
library(rhdf5)
library("DESeq2")
library(apeglm)
library(ashr)
library(stringr)
```

Parameters that will need changing per environment
```{r}
boot_base_dir <- "/home/humebc/projects/boote"

# Full path to the meta info file
meta_info_path <- file.path(boot_base_dir, "deseq_analysis/boote_meta_info.csv")

# The directory containing the host and zooxs directories that
# contain the by sample kallisto results
kallisto_base_dir <- file.path(boot_base_dir, "kallisto_quant_results")

# Directory where the results will be saved
base_output_dir <- file.path(boot_base_dir, "deseq_analysis", "deseq_results")

# A random selection of an Acropora and a Montipora sample to get the gene
# sets from
example_abund_table_path_acropora <- file.path(kallisto_base_dir, "Acropora", "AD1", "abundance.tsv")
example_abund_table_path_montipora <- file.path(kallisto_base_dir, "Montipora", "MD1", "abundance.tsv")

# The shrinkage method to implement
shrinkage_method_string = "ashr"

# In addition we will calculate the svalues using the ash function
# The S-values are calculated using the log2FoldChange and lfcSE variables
# or the non-shrinkage results
# They are appended to the results df before output.
```


Get the meta info
```{r}
meta_info <- read.csv(meta_info_path)
meta_info$Treatment <- as.factor(meta_info$Treatment)
meta_info$Species <- as.factor(unlist(lapply(str_split(meta_info$Species, pattern=" "), '[[', 1)))
# The sample names given do not exactly match the sample names as in the seq files
# so pull out the sample names from the seq files instead
meta_info$SampleName = str_extract(meta_info$Raw_read_R, "[:alpha:]+[:digit:]+")
```

Get the list of files for the given experimental setup
```{r}
get_file_list = function(submeta){
  files <- file.path(kallisto_base_dir, target_species, submeta$SampleName, "abundance.h5")
  names(files) <- submeta$SampleName
  return(files)
}
```

Make tx2gene df
```{r}
make_tx2gene_df = function(){
  # Create the tx2gene files that maps transcript to gene
  # In our case we are already working with gene so we will
  # create a dummy df that maps gene to gene
  # We will get the gene list from a random abundance table
  if (target_species == "Acropora"){
    example_table = read.table(example_abund_table_path_acropora, header = T)
    tx_list = example_table$target_id
    gene_list = str_extract(example_table$target_id, "aint_s[:digit:]+\\.g[:digit:]+")
  }else if(target_species == "Montipora"){
    example_table = read.table(example_abund_table_path_montipora, header = T)
    tx_list = example_table$target_id
    gene_list = str_extract(example_table$target_id, "TR[:digit:]+[:symbol:]{1}c[:digit:]+_g[:digit:]+")
  } else{
    stop(paste0("Unreconised value for target_species: ", target_species, "."))
  }
  return(data.frame("TXNAME"=tx_list, "GENEID"=gene_list))
}    
    
```

Output the CBASS.temperature contrasts (30vs33, 30vs36, 30vs39)
within the specified Timepoint and LT as DESEQ results objects
to a directory that is constructed from the 
```{r}
output_results = function(dds){
    pairwise_combinations = combn(c("Normoxy", "Deoxy", "Time_zero"), 2)
    for (i in 1:3){
      condition_one = pairwise_combinations[1,i]
      condition_two = pairwise_combinations[2,i]
      if (perform_shrinkage){
        # See here for further details https://rdrr.io/bioc/DESeq2/man/lfcShrink.html
        # In the PNAS paper that chris sent me:
        # https://onlinelibrary.wiley.com/doi/full/10.1111/mec.16064
        # that I was supposed to be following as the methods they used the S-value
        # to determine significant genes. The lfcShrink output does not give
        # you the S-value output. As such I will append this to the
        # res data.frame before output
        # The S-values are calculated using the log2FoldChange and lfcSE variables
        # or the non-shrinkage results
        results_no_shrinkage = results(dds, contrast=c("Treatment", condition_one, condition_two))
        ash_results = ash(results_no_shrinkage$log2FoldChange, results_no_shrinkage$lfcSE)
        results_shrinkage = lfcShrink(dds, contrast=c("Treatment", condition_one, condition_two), type=shrinkage_method_string)
        results_shrinkage$svalue = ash_results$result$svalue
        save_path_csv = file.path(base_output_dir, target_species, "shrinkage", paste0(condition_one, "_", condition_two, "shrinkage", ".csv"))
        save_path_obj = file.path(base_output_dir, target_species, "shrinkage", paste0(condition_one, "_", condition_two, "shrinkage", ".rds"))
        if (!dir.exists(file.path(base_output_dir, target_species, "shrinkage"))) {dir.create(file.path(base_output_dir, target_species, "shrinkage"), recursive = T)}
        write.csv(as.data.frame(results_shrinkage), file=save_path_csv)
        saveRDS(results_shrinkage, file=save_path_obj)
      }else{
        save_path_csv = file.path(base_output_dir, target_species, "no_shrinkage", paste0(condition_one, "_", condition_two, "no_shrinkage", ".csv"))
        save_path_obj = file.path(base_output_dir, target_species, "no_shrinkage", paste0(condition_one, "_", condition_two, "no_shrinkage", ".rds"))
        res = results(dds, contrast=c("Treatment", condition_one, condition_two))
        if (!dir.exists(file.path(base_output_dir, target_species, "no_shrinkage"))) {dir.create(file.path(base_output_dir, target_species, "no_shrinkage"), recursive = T)}
        write.csv(as.data.frame(res), file=save_path_csv)
        saveRDS(res, file=save_path_obj)
      }
    }
}
```


Currently we have been asked to do all pairwise contrast of the treatments
within each of the species.
```{r}
for (target_species in c("Acropora", "Montipora")){
  for (perform_shrinkage in c(TRUE, FALSE)){
    meta_info_sub = meta_info[meta_info$Species == target_species, ]
    files = get_file_list(meta_info_sub)
    tx2gene = make_tx2gene_df()
    txi = tximport(files, type = "kallisto", tx2gene = tx2gene)
    dds = DESeqDataSetFromTximport(txi, colData=meta_info_sub, design = ~ Treatment)
    dds = DESeq(dds)
    # Output the results for Normoxy vs Deoxy, Normoxy vs Trime_zero, Deoxy vs Time_zero
    output_results(dds=dds)
  }
}

```




